---
title: "Reproducible Research: Peer Assessment 2"
author: "Dr. B"
date: "Thursday, November 20, 2014"
output: html_document
---
Impact of Severe Weather Events on Public Health and the Economy

###Synonpsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
```{r,warning=FALSE,message=FALSE}
##Use my standard openning including call function
source('C:/Users/bryan_000/Documents/GitHub/MyWork/StdOpen.R')
```

###Data Processing 

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size.  The events in the database start in the year 1950 and end in November 2011. The dataset is downloaded and unzipped.
```{r,warning=FALSE,message=FALSE}
##Set destination file for dowload 
datafile <-paste(datadir,"stormData.csv.bz2",sep = "")

##Check to see if file exists, if not download it
if (!file.exists(datafile))
       {
                data.url <- 'http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
                download.file(data.url, datafile) 
                downloadedDataDate <- date()
       }

##Read and process the data
data <- read.csv(bzfile(datafile),header=TRUE, stringsAsFactors = FALSE)
names(data) <- tolower(names(data))     ##make all column names lowercase
names(data) <- gsub("_","",names(data)) ##remove underscores

##Add a year column if it does not exist
if (dim(data)[2] == 37) 
        {
         data$year <- as.numeric(format(as.Date(data$bgndate, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
        }

##Process the property damage
data$propdmgexp <- as.character(data$propdmgexp)
data$propdmgexp[toupper(data$propdmgexp) == 'H'] <- "2"
data$propdmgexp[toupper(data$propdmgexp) == 'K'] <- "3"
data$propdmgexp[toupper(data$propdmgexp) == 'M'] <- "6"
data$propdmgexp[toupper(data$propdmgexp) == 'B'] <- "9"
data$propdmgexp <- as.numeric(data$propdmgexp)
data$propdmgexp[is.na(data$propdmgexp)] <- 0
data$totalpropdmg <- data$propdmg * 10^data$propdmgexp

##Process the crop damage
data$cropdmgexp <- as.character(data$cropdmgexp)
data$cropdmgexp[toupper(data$cropdmgexp) == 'H'] <- "2"
data$cropdmgexp[toupper(data$cropdmgexp) == 'K'] <- "3"
data$cropdmgexp[toupper(data$cropdmgexp) == 'M'] <- "6"
data$cropdmgexp[toupper(data$cropdmgexp) == 'B'] <- "9"
data$cropdmgexp <- as.numeric(data$cropdmgexp)
data$cropdmgexp[is.na(data$cropdmgexp)] <- 0
data$totalcropdmg <- data$cropdmg * 10^data$cropdmgexp

##Set the years in the print title
maintitle=paste0("For the years ",min(data$year)," to ", max(data$year))
```

There are a total of `r nrow(data)` observations in this data set with `r dim(data)[2]` columns.  

###Results

####What types of events are the most harmful with respect to population health?

First, I performed an aggregate function to obtain the sum of all fatalities caused by an specific event. From that, I ordered the data in a decreasing order, took the top 10, and prepared a barplot.  Next, I performed an aggregate function to obtain the sum of all injuries caused by an specific event. From that, I ordered the data in a decreasing order, took the top 10, and prepared a barplot.
```{r,warning=FALSE,message=FALSE}
##Sum the fatalities and plot the top 10
deaths <- aggregate(data$fatalities, by = list(data$evtype), "sum")
names(deaths) <- c("event", "fatalities")
deaths <- deaths[order(-deaths$fatalities), ][1:10, ]

par(mfrow=c(2,2))
barplot(deaths$fatalities, names.arg = deaths$event, col = 'yellow',main = 'Top 10 Weather Events \n for Fatalities', ylab = 'Number of Fatalities')

##Sum the injuries and plot the top 10
injuries <- aggregate(data$injuries, by = list(data$evtype), "sum")
names(injuries) <- c("event", "injuries")
injuries <- injuries[order(-injuries$injuries), ][1:10, ]

barplot(injuries$injuries, names.arg = injuries$event, col = 'green', main = 'Top 10 Weather Events \n for Injuries', ylab = 'Number of Injuries')

## Put a title over both plots
mtext(maintitle, side = 3, line = -1, outer = TRUE)
```

####What types of events ave the greatest economic consequences?

First, I performed an aggregate function to obtain the sum of all the property damage caused by an specific event. From that, I ordered the data in a decreasing order, took the top 10, and prepared a barplot.  Next, I performed an aggregate function to obtain the sum of all the crop damage caused by an specific event. From that, I ordered the data in a decreasing order, took the top 10, and prepared a barplot.
```{r, warning=FALSE, message=FALSE}
##Calculate property damage and plot
sumpropdmg <- aggregate(data$totalpropdmg, by = list(data$evtype), "sum")
names(sumpropdmg) <- c("event", "cost")
sumpropdmg <- sumpropdmg[order(-sumpropdmg$cost), ][1:10, ]

par(mfrow=c(2,2))
barplot(sumpropdmg$cost, names.arg = sumpropdmg$event, col = 'blue', main = 'Top 10 Weather Events \n for Property Damage', ylab = 'Cost of Damage')

##Calculate crop damage and plot
sumcropdmg <- aggregate(data$totalcropdmg, by = list(data$evtype), "sum")
names(sumcropdmg) <- c("event", "cost")
sumcropdmg <- sumcropdmg[order(-sumcropdmg$cost), ][1:10, ]

barplot(sumcropdmg$cost, names.arg = sumpropdmg$event, col = 'red', main = 'Top 10 Weather Events \n for Crop Damage', ylab = 'Cost of Damage')

##Place a title overr both plots
mtext(maintitle, side = 3, line = -1, outer = TRUE)
```

###Conclusion

The top 10 harmful weather events to public health and the economy are as follows:
```{r}
summarydf <-data.frame(deathevent=deaths$event, deathnbr=deaths$fatalities, injevent=injuries$event, injnbr=injuries$injuries, propdmgevent=sumpropdmg$event,propcost=sumpropdmg$cost,cropdmgevent=sumcropdmg$event,cropcost=sumcropdmg$cost)
summarydf
```
